Parameter Recovery Validation

### 2025-05-28 nbs/concepts/parameter-recovery-validation/index.qmd
docs: add section on piecewise constant activation function to parameter recovery validation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
- **Dynamics**: Same dimensionless analytical solutions
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
- **Prior**: Gamma priors for time scales, Normal for relative positions
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{max}^*$, $t_{loc}$, $t_{scale}$, $\tilde{t}_j$)
- **Prior**: Gamma priors for time scales, Normal for relative positions
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

# hyperparameters for hierarchical time structure (moved further left)
pgm.add_node("mu_T", r"$\mu_{T}$", 5.5, 6, fixed=True)
pgm.add_node("sigma_T", r"$\sigma_{T}^2$", 6, 6, fixed=True)
pgm.add_node("mu_t_loc", r"$\mu_{t_{loc}}$", 6.5, 6, fixed=True)
pgm.add_node("sigma_t_loc", r"$\sigma_{t_{loc}}^2$", 7, 6, fixed=True)
pgm.add_node("mu_t_scale", r"$\mu_{t_{scl}}$", 7.5, 6, fixed=True)
pgm.add_node("sigma_t_scale", r"$\sigma_{t_{scl}}^2$", 8, 6, fixed=True)

@@

# hierarchical time structure edges
pgm.add_edge("mu_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("sigma_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("mu_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("sigma_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("mu_t_scale", "t_scale", plot_params=edge_params)
pgm.add_edge("sigma_t_scale", "t_scale", plot_params=edge_params)

@@

**Hierarchical priors for latent time coordinates:**
\begin{align}
    T_{max}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \label{eq-T-max-prior} \\
    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \label{eq-t-loc-prior} \\
    t_{scale} &\sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}}) \label{eq-t-scale-prior} \\
    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scale}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-t-tilde-prior} \\
    t^*_j &= T_{max}^* \times \tilde{t}_j \quad \text{(deterministic transformation)} \label{eq-t-star-transform}
\end{align}

$\tilde{t}_j$ represents the **relative temporal position** of cell $j$ (which can be negative for cells before the process start or greater than 1 for cells after the typical process duration), and $t^*_j$ represents the **actual dimensionless time coordinate** obtained by scaling the relative position by the global time scale $T_{max}^*$.

@@

**$T_{max}^*$ (Global dimensionless time scale)**

- **Definition**: Global scaling factor for dimensionless time coordinates
- **Units**: Dimensionless (in units of characteristic splicing time)
- **Interpretation**: Characteristic temporal extent of the biological process
- **Biological meaning**: Represents the typical time scale over which the biological process unfolds
- **Model role**: Multiplicative scaling factor in $t^*_j = T_{max}^* \times \tilde{t}_j$
- **Prior**: $T_{max}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$ ensures positive values with flexible tail behavior
- **Example**: $T_{max}^* = 5$ means the process typically spans ~5 characteristic splicing times

@@

**$t_{scale}$ (Population relative time scale)**

- **Definition**: Population-level scale parameter for relative time distribution
- **Units**: Dimensionless (relative time units)
- **Interpretation**: How much relative temporal variation exists across cells
- **Biological meaning**: Degree of temporal heterogeneity in the cell population
- **Prior**: $t_{scale} \sim \text{Gamma}(\alpha_{t_{scale}}, \beta_{t_{scale}})$ ensures positive scale with flexible shape
- **Statistical role**: Controls the amount of shrinkage toward the population mean
- **Example**: $t_{scale} = 0.2$ indicates relatively tight temporal clustering around $t_{loc}$

@@

- **Definition**: Latent relative time coordinate for cell $j$
- **Units**: Dimensionless (relative time units)
- **Interpretation**: Cell's position relative to the population temporal distribution
- **Biological meaning**: Where this cell sits in the relative temporal progression
- **Prior**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scale}^2)$ allows negative values (before process) and values >1 (after process)
- **Flexibility**: Can represent cells outside the typical process timeline
- **Example**: $\tilde{t}_j = -0.2$ means this cell is before the typical process start; $\tilde{t}_j = 1.3$ means after typical completion

@@

- **Definition**: Actual dimensionless time coordinate for cell $j$
- **Units**: Dimensionless (in units of characteristic splicing time)
- **Interpretation**: How many splicing events have elapsed for this specific cell
- **Biological meaning**: Cell's absolute position along the temporal trajectory of the biological process
- **Transformation**: $t^*_j = T_{max}^* \times \tilde{t}_j$ (deterministic scaling)
- **Example**: If $T_{max}^* = 5$ and $\tilde{t}_j = 0.6$, then $t^*_j = 3.0$ characteristic splicing times

@@

- **Improved identifiability**: Hierarchical structure provides regularization without redundant parameters
- **Borrowing strength**: Information shared across cells improves parameter estimation
- **Uncertainty quantification**: Full posterior distributions for all time parameters
- **Robustness**: Less sensitive to outlier cells or sparse data
- **No truncation constraints**: Avoids computational complexity of truncated distributions
- **Clear parameter roles**: $T_{max}^*$ controls global scale, $t_{loc}$ and $t_{scale}$ control population distribution

@@

- **Dimensionless time**: Universal interpretation across experimental conditions
- **Population heterogeneity**: Captures natural variation in biological timing
- **Individual trajectories**: Each cell has its own temporal position
- **Flexible temporal range**: Cells can be before ($\tilde{t}_j < 0$) or after ($\tilde{t}_j > 1$) typical process timeline
- **Natural biological variation**: No artificial hard boundaries on temporal positions

@@

- **Standard distributions**: Gamma and Normal distributions are computationally efficient
- **No truncation**: Avoids complex truncated distribution sampling
- **Analytical solutions**: Fast likelihood evaluation for validation
- **Hierarchical sampling**: Efficient MCMC mixing through population-level parameters
- **Multiplicative structure**: Simple deterministic transformation $t^*_j = T_{max}^* \times \tilde{t}_j$

## Validation Study 3: Dimensionless simulated dynamics with observed time

*To be implemented: Replace analytical solutions with numerical integration using torchode while maintaining the dimensionless parameterization.*

## Validation Study 4: Dimensionless simulated dynamics with latent time

*To be implemented: Full complexity model with numerical integration and latent time coordinates, representing the complete PyroVelocity modular implementation.*

>>>>

- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
- **Dynamics**: Same dimensionless analytical solutions
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{M}^*$, $t_{loc}$, $t_{scl}$)
- **Prior**: Gamma priors for time scales, Normal for relative positions
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{M}^*$, $t_{loc}$, $t_{scl}$)
- **Prior**: Gamma priors for time scales, Normal for relative positions
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

# hyperparameters for hierarchical time structure
pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)
pgm.add_node("alpha_t_loc", r"$\alpha_{t_{loc}}$", 6.5, 6, fixed=True)
pgm.add_node("beta_t_loc", r"$\beta_{t_{loc}}$", 7, 6, fixed=True)
pgm.add_node("alpha_t_scale", r"$\alpha_{t_{scl}}$", 7.5, 6, fixed=True)
pgm.add_node("beta_t_scale", r"$\beta_{t_{scl}}$", 8, 6, fixed=True)

@@

# hierarchical time structure edges
pgm.add_edge("alpha_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("beta_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("alpha_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("beta_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("alpha_t_scale", "t_scale", plot_params=edge_params)
pgm.add_edge("beta_t_scale", "t_scale", plot_params=edge_params)

@@

**Hierarchical priors for latent time coordinates:**
\begin{align}
    T_{M}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \label{eq-T-M-prior} \\
    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \label{eq-t-loc-prior} \\
    t_{scl} &\sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}}) \label{eq-t-scl-prior} \\
    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scl}^2) \quad \text{for } j \in \{1, \ldots, N\} \label{eq-t-tilde-prior} \\
    t^*_j &= T_{M}^* \times \max(\tilde{t}_j, \epsilon) \quad \text{(clipped transformation)} \label{eq-t-star-transform}
\end{align}

$\tilde{t}_j$ represents the **relative temporal position** of cell $j$ within the population distribution, and $t^*_j$ represents the **actual dimensionless time coordinate**. The clipping operation $\max(\tilde{t}_j, \epsilon)$ with $\epsilon = 10^{-5}$ ensures that all time coordinates remain positive, maintaining biological realism (cells cannot be "before" the process) and mathematical consistency with the analytical solutions. While $\tilde{t}_j$ is not explicitly shown in the PGM for space constraints, the clipped transformation is implicitly included in the $t^*_j$ node.

@@

**$T_{M}^*$ (Global dimensionless time scale)**

- **Definition**: Global scaling factor for dimensionless time coordinates
- **Units**: Dimensionless (in units of characteristic splicing time)
- **Interpretation**: Characteristic temporal extent of the biological process
- **Biological meaning**: Represents the typical time scale over which the biological process unfolds
- **Model role**: Multiplicative scaling factor in $t^*_j \sim T_{M}^* \times \text{Normal}(t_{loc}, t_{scl}^2)$
- **Prior**: $T_{M}^* \sim \text{Gamma}(\alpha_{T}, \beta_{T})$ ensures positive values with flexible tail behavior
- **Example**: $T_{M}^* = 5$ means the process typically spans ~5 characteristic splicing times

@@

**$t_{scl}$ (Population relative time scale)**

- **Definition**: Population-level scale parameter for relative time distribution
- **Units**: Dimensionless (relative time units)
- **Interpretation**: How much relative temporal variation exists across cells
- **Biological meaning**: Degree of temporal heterogeneity in the cell population
- **Prior**: $t_{scl} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})$ ensures positive scale with flexible shape
- **Statistical role**: Controls the amount of shrinkage toward the population mean
- **Example**: $t_{scl} = 0.2$ indicates relatively tight temporal clustering around $t_{loc}$

@@

- **Definition**: Latent relative time coordinate for cell $j$ within the population distribution
- **Units**: Dimensionless (relative time units)
- **Interpretation**: Cell's position relative to the population temporal distribution
- **Biological meaning**: Where this cell sits in the relative temporal progression
- **Prior**: $\tilde{t}_j \sim \text{Normal}(t_{loc}, t_{scl}^2)$ allows natural population variation
- **Range**: Can be negative (before typical process) or greater than 1 (after typical process)
- **Example**: $\tilde{t}_j = -0.2$ means this cell would be before the typical process start; $\tilde{t}_j = 1.3$ means after typical completion

@@

- **Definition**: Actual dimensionless time coordinate for cell $j$ after clipping
- **Units**: Dimensionless (in units of characteristic splicing time)
- **Interpretation**: How many splicing events have elapsed for this specific cell
- **Biological meaning**: Cell's absolute position along the temporal trajectory of the biological process
- **Transformation**: $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$ with $\epsilon = 10^{-5}$
- **Constraint**: Always positive, ensuring mathematical consistency with analytical solutions
- **Example**: If $T_{M}^* = 5$, $\tilde{t}_j = 0.6$, then $t^*_j = 3.0$ characteristic splicing times; if $\tilde{t}_j = -0.1$, then $t^*_j = 5 \times 10^{-5}$ (clipped)

@@

- **Improved identifiability**: Hierarchical structure provides regularization without redundant parameters
- **Borrowing strength**: Information shared across cells improves parameter estimation
- **Uncertainty quantification**: Full posterior distributions for all time parameters
- **Robustness**: Less sensitive to outlier cells or sparse data
- **No truncation constraints**: Avoids computational complexity of truncated distributions
- **Clear parameter roles**: $T_{M}^*$ controls global scale, $t_{loc}$ and $t_{scl}$ control population distribution

@@

- **Dimensionless time**: Universal interpretation across experimental conditions
- **Population heterogeneity**: Captures natural variation in biological timing
- **Individual trajectories**: Each cell has its own temporal position
- **Flexible temporal range**: $\tilde{t}_j$ can represent cells before or after the typical process timeline
- **Biological realism**: Clipping ensures cells cannot be "before" the process start ($t^*_j \geq \epsilon$)
- **Mathematical consistency**: Positive time coordinates maintain validity of analytical solutions

@@

- **Standard distributions**: Gamma and Normal distributions are computationally efficient
- **No truncation**: Avoids complex truncated distribution sampling
- **Simple clipping**: $\max(\tilde{t}_j, \epsilon)$ operation is computationally trivial
- **Analytical solutions**: Fast likelihood evaluation for validation
- **Hierarchical sampling**: Efficient MCMC mixing through population-level parameters
- **Deterministic transformation**: $t^*_j = T_{M}^* \times \max(\tilde{t}_j, \epsilon)$ is simple and fast

## Validation Study 3: Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation

This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving the identifiability issues discovered in Study 2. The key innovation is the simultaneous introduction of piecewise activation dynamics and steady-state initial conditions, which together create biologically meaningful dynamics while ensuring proper parameter identifiability.

The model extends Study 2 by replacing the constant transcription rate $\alpha^* = 1$ with a **piecewise constant transcription function** $\alpha^*_i(t^*)$ that varies over time for each gene. Initial conditions are fixed to the steady-state values associated with the "off" phase, eliminating the redundant parameterization between state and temporal initial conditions.

### Piecewise Transcription Rate Function

For each gene $i$, the dimensionless transcription rate follows a three-phase piecewise constant function:

$$\alpha^*_i(t^*) := \begin{cases}
\alpha^*_{\text{off}i} & \text{if } 0 \leq t^* < t^*_{0,\text{on}i}, \\
\alpha^*_{\text{on}i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
\alpha^*_{\text{off}i} & \text{if } t^*_{0,\text{on}i} + \delta^*_i \leq t^*
\end{cases}$$

where:
- $\alpha^*_{\text{off}i} = \alpha_{\text{off}i}/\beta_i$ is the dimensionless basal transcription rate
- $\alpha^*_{\text{on}i} = \alpha_{\text{on}i}/\beta_i$ is the dimensionless active transcription rate
- $t^*_{0,\text{on}i} = \beta_i \cdot t_{0,\text{on}i}$ is the dimensionless activation onset time
- $\delta^*_i = \beta_i \cdot \delta_i$ is the dimensionless activation duration

### Steady-State Initial Conditions

To resolve identifiability, initial conditions are fixed to the steady-state values corresponding to the "off" phase:

$$u^*_{0i} = \alpha^*_{\text{off}i}, \quad s^*_{0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$

This eliminates the need to infer $u^*_{0i}$ and $s^*_{0i}$ as separate parameters while providing equivalent modeling flexibility through the $\alpha^*_{\text{off}i}$ parameter.

### Analytical Solutions for Each Phase

The dimensionless system with piecewise activation becomes:

$$\begin{align}
\frac{du^{\ast}}{dt^{\ast}} &= \alpha^*_i(t^*) - u^{\ast} \\
\frac{ds^{\ast}}{dt^{\ast}} &= u^{\ast} - \gamma^{\ast}_i s^{\ast}
\end{align}$$

We solve this system analytically for each of the three phases:

#### **Phase 1: Off State** ($0 \leq t^* < t^*_{0,\text{on}i}$)

With constant transcription rate $\alpha^*_{\text{off}i}$ and initial conditions at steady state, the system remains at equilibrium:

$$\begin{align}
u^*_{ij}(t^*) &= \alpha^*_{\text{off}i} \\
s^*_{ij}(t^*) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}
\end{align}$$

#### **Phase 2: On State** ($t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i$)

Let $\tau_{\text{on}} = t^* - t^*_{0,\text{on}i}$ be the time since activation onset. The system transitions from the off-state steady state toward the on-state steady state.

**Initial conditions for Phase 2:**
$$u^*_{\text{on},0i} = \alpha^*_{\text{off}i}, \quad s^*_{\text{on},0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$

**Analytical solutions:**

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
s^*_{ij}(\tau_{\text{on}}) &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{on}}} + \xi_{\text{on}i} e^{-\tau_{\text{on}}}
\end{align}$$

where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$.

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
s^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \tau_{\text{on}} e^{-\tau_{\text{on}}}
\end{align}$$

#### **Phase 3: Return to Off State** ($t^* \geq t^*_{0,\text{on}i} + \delta^*_i$)

Let $\tau_{\text{off}} = t^* - (t^*_{0,\text{on}i} + \delta^*_i)$ be the time since returning to the off state. The system transitions from the end-of-phase-2 state back toward the off-state steady state.

**Initial conditions for Phase 3:**

The initial conditions for Phase 3 are the endpoint values from Phase 2, computed explicitly as:

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
s^*_{\text{off},0i} &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \delta^*_i} + \xi_{\text{on}i} e^{-\delta^*_i}
\end{align}$$

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
s^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \delta^*_i e^{-\delta^*_i}
\end{align}$$

where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ for the $\gamma^*_i \neq 1$ case.

**Analytical solutions:**

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
s^*_{ij}(\tau_{\text{off}}) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i} + \left(s^*_{\text{off},0i} - \xi_{\text{off}i} - \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{off}}} + \xi_{\text{off}i} e^{-\tau_{\text{off}}}
\end{align}$$

where $\xi_{\text{off}i} = \frac{u^*_{\text{off},0i} - \alpha^*_{\text{off}i}}{\gamma^*_i - 1}$.

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
s^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (s^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) \tau_{\text{off}} e^{-\tau_{\text{off}}}
\end{align}$$

### Gene Expression Pattern Capabilities

This piecewise framework enables representation of diverse gene expression patterns:

1. **Activation patterns**: Low $\alpha^*_{\text{off}i}$, early $t^*_{0,\text{on}i}$, high $\alpha^*_{\text{on}i}$
2. **Decay-only patterns**: High $\alpha^*_{\text{off}i}$, $t^*_{0,\text{on}i} > T^*_M$ (onset beyond observation window)
3. **Transient activation**: Early $t^*_{0,\text{on}i}$, short $\delta^*_i$, moderate $\alpha^*_{\text{on}i}$
4. **Sustained activation**: Early $t^*_{0,\text{on}i}$, long $\delta^*_i$, high $\alpha^*_{\text{on}i}$

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure. For visual clarity, we use $\theta_i$ to represent the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$ and $\delta_i$ to represent the activation duration parameter.

```{python}
#| label: fig-piecewise-activation-pgm
#| code-fold: true
#| fig-cap: Probabilistic graphical model for dimensionless analytical dynamics with hierarchical latent time coordinates and piecewise activation.

# hyperparameters for gene-specific parameters
pgm.add_node("mu_theta", r"$\mu_{\theta}$", 0.5, 6, fixed=True)
pgm.add_node("sigma_theta", r"$\sigma_{\theta}^2$", 1.5, 6, fixed=True)
pgm.add_node("mu_delta", r"$\mu_{\delta}$", 2.5, 6, fixed=True)
pgm.add_node("sigma_delta", r"$\sigma_{\delta}^2$", 3.5, 6, fixed=True)
pgm.add_node("mu_U0", r"$\mu_{U_0}$", 4.5, 6, fixed=True)
pgm.add_node("sigma_U0", r"$\sigma_{U_0}^2$", 5, 6, fixed=True)

# hyperparameters for hierarchical time structure
pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)
pgm.add_node("alpha_t_loc", r"$\alpha_{t_{loc}}$", 6.5, 6, fixed=True)
pgm.add_node("beta_t_loc", r"$\beta_{t_{loc}}$", 7, 6, fixed=True)
pgm.add_node("alpha_t_scale", r"$\alpha_{t_{scl}}$", 7.5, 6, fixed=True)
pgm.add_node("beta_t_scale", r"$\beta_{t_{scl}}$", 8, 6, fixed=True)

# latent variables for gene-specific parameters
pgm.add_node("theta_i", r"$\theta_i$", 1, 5)
pgm.add_node("delta_i", r"$\delta_i$", 3, 5)
pgm.add_node("U_0i", r"$U_{0i}$", 4.5, 5)

# gene-specific parameter edges
pgm.add_edge("mu_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("sigma_theta", "theta_i", plot_params=edge_params)
pgm.add_edge("mu_delta", "delta_i", plot_params=edge_params)
pgm.add_edge("sigma_delta", "delta_i", plot_params=edge_params)
pgm.add_edge("mu_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("sigma_U0", "U_0i", plot_params=edge_params)
pgm.add_edge("mu_lambda", "lambda_j", plot_params=edge_params)
pgm.add_edge("sigma_lambda", "lambda_j", plot_params=edge_params)

# hierarchical time structure edges
pgm.add_edge("alpha_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("beta_T", "T_max_star", plot_params=edge_params)
pgm.add_edge("alpha_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("beta_t_loc", "t_loc", plot_params=edge_params)
pgm.add_edge("alpha_t_scale", "t_scale", plot_params=edge_params)
pgm.add_edge("beta_t_scale", "t_scale", plot_params=edge_params)

# dynamics edges
pgm.add_edge("theta_i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("theta_i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("delta_i", "u_star_ij", plot_params=edge_params)
pgm.add_edge("delta_i", "s_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "u_star_ij", plot_params=edge_params)
pgm.add_edge("t_star_j", "s_star_ij", plot_params=edge_params)

The model combines the hierarchical latent time structure from Study 2 with piecewise constant transcription rates and steady-state initial conditions to resolve identifiability while enabling rich gene expression dynamics.

**Notation mapping between PGM and mathematical description:**

In the PGM above, we use simplified notation for visual clarity:
- $\theta_i$ represents the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$
- $\delta_i$ represents the activation duration parameter $\delta^*_i$ (waiting time from onset to deactivation)
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\delta}, \sigma_{\delta}^2$ represent the hyperparameters for $\delta_i$

**Detailed priors for piecewise activation parameters:**

The priors are designed to encode biologically reasonable parameter ranges while maintaining sufficient flexibility for diverse gene expression patterns:

\begin{align}
    \gamma^*_i &\sim \text{LogNormal}(\log(1.0), 0.5^2) \quad \text{(relative degradation rate)} \\
    \alpha^*_{\text{off}i} &\sim \text{LogNormal}(\log(0.1), 0.5^2) \quad \text{(basal transcription rate)} \\
    \alpha^*_{\text{on}i} &\sim \text{LogNormal}(\log(2.0), 0.5^2) \quad \text{(active transcription rate)} \\
    t^*_{0,\text{on}i} &\sim \text{LogNormal}(\log(0.3), 0.3^2) \quad \text{(activation onset time)} \\
    \delta^*_i &\sim \text{LogNormal}(\log(0.5), 0.3^2) \quad \text{(activation duration)}
\end{align}

**Biological interpretation of prior choices:**

- **$\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$**: Centers relative degradation around 1 (balanced splicing/degradation), with 95% of values between ~0.37 and ~2.7. This covers genes where mRNA is more stable than splicing time ($\gamma^* < 1$) to genes with rapid degradation ($\gamma^* > 1$).

- **$\alpha^*_{\text{off}i} \sim \text{LogNormal}(\log(0.1), 0.5^2)$**: Centers basal transcription at 10% of characteristic rate, with 95% between ~0.037 and ~0.27. This represents low but non-zero basal expression typical of repressed genes.

- **$\alpha^*_{\text{on}i} \sim \text{LogNormal}(\log(2.0), 0.5^2)$**: Centers active transcription at 2× characteristic rate, with 95% between ~0.74 and ~5.4. This ensures activation represents meaningful upregulation while avoiding extreme values.

- **$t^*_{0,\text{on}i} \sim \text{LogNormal}(\log(0.3), 0.3^2)$**: Centers activation onset at 30% through the process timeline, with 95% between ~0.17 and ~0.53. This favors early activation while allowing variation in timing.

- **$\delta^*_i \sim \text{LogNormal}(\log(0.5), 0.3^2)$**: Centers activation duration at 50% of the process timeline, with 95% between ~0.28 and ~0.89. This allows both transient and sustained activation patterns.

**Priors for other gene-specific parameters (unchanged from Study 2):**
\begin{align}
    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
\end{align}

**Hierarchical priors for latent time coordinates (unchanged from Study 2):**
\begin{align}
    T_{M}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \\
    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \\
    t_{scl} &\sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}}) \\
    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scl}^2) \quad \text{for } j \in \{1, \ldots, N\} \\
    t^*_j &= T_{M}^* \times \max(\tilde{t}_j, \epsilon) \quad \text{(clipped transformation)}
\end{align}

**Cell-specific capture efficiency (unchanged from Study 2):**
\begin{align}
    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
\end{align}

**Dimensionless dynamics with piecewise activation:**

The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$.

### Key Advantages of the Piecewise Framework

#### **Identifiability Resolution**

- **Eliminates redundant parameterization**: No separate inference of $u^*_{0i}, s^*_{0i}$ parameters
- **Biological realism**: Steady-state initial conditions are more plausible than arbitrary values
- **Clear parameter roles**: Each piecewise parameter has distinct biological interpretation

#### **Rich Expression Pattern Representation**

- **Activation genes**: Transition from low to high expression
- **Decay genes**: Start high, remain high (no activation within observation window)
- **Transient genes**: Brief activation followed by return to baseline
- **Sustained genes**: Long-duration activation

#### **Analytical Tractability**

- **Fast likelihood evaluation**: All phases solved analytically
- **Efficient parameter recovery validation**: No numerical integration required
- **Clear mathematical structure**: Each phase has explicit closed-form solution

- **Dimensionless parameters**: Universal interpretation across experimental conditions
- **Temporal coordination**: Hierarchical time structure captures population-level timing
- **Gene-specific dynamics**: Each gene can follow distinct activation pattern

### Implementation Considerations

#### **Phase Detection Logic**

For implementation, each cell-gene pair $(j,i)$ must be assigned to the correct phase based on the cell's latent time $t^*_j$ relative to gene $i$'s activation parameters:

```python
def determine_phase(t_star_j, t_on_i, delta_i):
    """Determine which phase cell j is in for gene i."""
    if t_star_j < t_on_i:
        return "off_phase_1"  # Phase 1: before activation
    elif t_star_j < t_on_i + delta_i:
        return "on_phase"     # Phase 2: during activation
    else:
        return "off_phase_3"  # Phase 3: after activation
```

#### **Numerical Stability Considerations**

**Special case handling for $\gamma^*_i \approx 1$:**

When $\gamma^*_i$ is very close to 1 (e.g., $|\gamma^*_i - 1| < 10^{-6}$), the helper variable $\xi_i = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ becomes numerically unstable. Implementation should:

1. **Use threshold switching**: If $|\gamma^*_i - 1| < \epsilon$ (e.g., $\epsilon = 10^{-6}$), use the $\gamma^*_i = 1$ analytical solutions
2. **L'Hôpital's rule**: For intermediate values, the limit as $\gamma^*_i \to 1$ gives the correct $t^* e^{-t^*}$ terms
3. **Robust computation**: Avoid direct division by $(\gamma^*_i - 1)$ when close to zero

**Parameter validation ranges:**

To ensure biological realism and numerical stability:
